---
title: "Vegetation Analysis InnDikes"
output: html_notebook
---

```{r}
library(here)
library(dplyr)
library(readr)
library(vegan)
library(ggplot2)
```

## Methods
Todo: write method section 

## Analysis: Vegetation Development over time 
### First year: compare seeded with early development
### A) Descriptive (NMDS)
### B) Model: is distance between seeded community (target) and established community affected by substrate depth or species pool?
### C) TODO: add reference vegetation data set 
```{r}
# first year

# include data
Inn_Dike_species <- read.csv2(
  here("data", "raw", "Inn_Dike_species.csv"),
  row.names = 1,     # use first column as rownames
  stringsAsFactors = FALSE,  # prevent automatic factor conversion
  check.names = FALSE       # preserve original column names
)

# Check
View(Inn_Dike_species)
```

## the data needs to be split into the two different species pools, otherwise the NMDS gets too little stress as species are occuring in too low frequency. 
```{r}
Inn_Dike_species <- as.data.frame(t(Inn_Dike_species))

Inn_Dike_species_22 <- subset(Inn_Dike_species, pool == "r22")

Inn_Dike_species_1A  <- subset(Inn_Dike_species, pool == "r1a")
```


```{r}
# split data into species composition and header
View(Inn_Dike_species_22)

Inn_Dike_species_22_head <- Inn_Dike_species_22[,c(1:6)]
View(Inn_Dike_species_22_head)

Inn_Dike_species_22_specs <- Inn_Dike_species_22[,-c(1:6)]
View(Inn_Dike_species_22_specs)

species_names_22 <- names(Inn_Dike_species_22_specs)
cols_original_22 <- rownames(Inn_Dike_species_22_specs)

# Convert all values to numeric
Inn_Dike_species_22_specs <- as.data.frame(lapply(Inn_Dike_species_22_specs, function(x) as.numeric(as.character(x))))
View(Inn_Dike_species_22_specs)

# Set species names as row names
rownames(Inn_Dike_species_22_specs) <- cols_original_22

Inn_Dike_species_22_specs <- as.data.frame(t(Inn_Dike_species_22_specs))

#----

View(Inn_Dike_species_1A)

Inn_Dike_species_1A_head <- Inn_Dike_species_1A[,c(1:6)]
View(Inn_Dike_species_1A_head)

Inn_Dike_species_1A_specs <- Inn_Dike_species_1A[,-c(1:6)]
View(Inn_Dike_species_1A_specs)

species_names_1A <- names(Inn_Dike_species_1A_specs)
cols_original_1A <- rownames(Inn_Dike_species_1A_specs)

# Convert all values to numeric
Inn_Dike_species_1A_specs <- as.data.frame(lapply(Inn_Dike_species_1A_specs, function(x) as.numeric(as.character(x))))
View(Inn_Dike_species_1A_specs)

# Set species names as row names
rownames(Inn_Dike_species_1A_specs) <- cols_original_1A

Inn_Dike_species_1A_specs <- as.data.frame(t(Inn_Dike_species_1A_specs))
```


# how frequent are species occuring? 
```{r}
Inn_Dike_species_22_specs$Vorkommen <- rowSums(Inn_Dike_species_22_specs > 0, na.rm = T)/ncol(Inn_Dike_species_22_specs)
hist(Inn_Dike_species_22_specs$Vorkommen, nclass = 20)

Inn_Dike_species_1A_specs$Vorkommen <- rowSums(Inn_Dike_species_1A_specs > 0, na.rm = T)/ncol(Inn_Dike_species_1A_specs)
hist(Inn_Dike_species_1A_specs$Vorkommen, nclass = 20)
```

# cut species that are not occuring (TODO later: cut species at certain treshold)
```{r}
Inn_Dike_species_22_specs <- subset(Inn_Dike_species_22_specs, Vorkommen > 0)
Inn_Dike_species_1A_specs <- subset(Inn_Dike_species_1A_specs, Vorkommen > 0)
```


# calculate ratio of each species for each plot - divide each value by sum per col
```{r}
# Step 1: compute column sums, ignoring NAs
col_sums_22 <- colSums(Inn_Dike_species_22_specs, na.rm = TRUE)
col_sums_1A <- colSums(Inn_Dike_species_1A_specs, na.rm = TRUE)

# Step 2: compute ratio of each value to its column sum (×100 for percentage)
Inn_Dike_species_22_ratio <- sweep(Inn_Dike_species_22_specs, 2, col_sums_22, FUN = "/")
Inn_Dike_species_1A_ratio <- sweep(Inn_Dike_species_1A_specs, 2, col_sums_1A, FUN = "/")

# View result
View(Inn_Dike_species_22_ratio)
View(Inn_Dike_species_1A_ratio)
```

```{r}
Inn_Dike_species_22_ratio <- as.data.frame(t(Inn_Dike_species_22_ratio[,-90])) # adjust col-number; double check, must be 96 (seeded) plus 116 (vegetation) 
Inn_Dike_species_1A_ratio <- as.data.frame(t(Inn_Dike_species_1A_ratio[,-c(1,97)]))
```

```{r}
any(is.na(Inn_Dike_species_22_ratio))
any(is.nan(as.matrix(Inn_Dike_species_22_ratio)))
any(is.infinite(as.matrix(Inn_Dike_species_22_ratio)))
```


```{r}
Inn_Dike_species_1A_ratio[is.na(Inn_Dike_species_1A_ratio)] <- 0
Inn_Dike_species_22_ratio[is.na(Inn_Dike_species_22_ratio)] <- 0
```

```{r}
View(Inn_Dike_species_1A_ratio)
```

```{r}
NMDS_Inn_1A <- metaMDS(Inn_Dike_species_1A_ratio[-35,], k=2, trymax=100)
```


```{r}
stressplot(NMDS_Inn_1A)
```

```{r}
NMDS_Inn_1A
```

```{r}
ordiplot(NMDS_Inn_1A)
ordiplot(NMDS_Inn_1A, type = 't')
```

# prepare header data
```{r}
View(Inn_Dike_species_1A_head)
Inn_Dike_species_1A_head <- as.data.frame(t(Inn_Dike_species_1A_head))
```

```{r}
# Extract NMDS scores for sites
scores_NMDS_1A <- as.data.frame(scores(NMDS_Inn_1A, display = "sites"))

# Add grouping variable (e.g., Management type)
scores_NMDS_1A$Group <- Inn_Dike_species_1A_head$timepoint # when header data is transposed correctly, timepoint is a col!

# Plot NMDS with ellipses for each group
ggplot(scores_NMDS_1A, aes(x = NMDS1, y = NMDS2, color = Group, fill = Group)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(type = "t", level = 0.95, alpha = 0.2, geom = "polygon") +
  theme_minimal(base_size = 14) +
  labs(title = "NMDS Plot with Group Ellipses",
       x = "NMDS1", y = "NMDS2") +
  theme(legend.position = "right")
```

```{r}
# Add grouping variable (e.g., Management type)
scores_NMDS_1A$Group2 <- interaction(Inn_Dike_species_1A_head$timepoint, Inn_Dike_species_1A_head$trait_constraint)

# Plot NMDS with ellipses for each group
ggplot(scores_NMDS_1A, aes(x = NMDS1, y = NMDS2, color = Group2, fill = Group2)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(type = "t", level = 0.95, alpha = 0.2, geom = "polygon") +
  theme_minimal(base_size = 14) +
  labs(title = "NMDS Plot with Group Ellipses",
       x = "NMDS1", y = "NMDS2") +
  theme(legend.position = "right")
```

## prepare and expand head; 
## tidy up todo: prepare one master header with all information; e.g. also the soil data etc. and use this from the beginning; 

```{r}
biomasse <- read_delim(
  here("data", "raw", "biomasse.csv"),
  delim = ";",
  escape_double = FALSE,
  trim_ws = TRUE
)
View(biomasse)
View(Inn_Dike_species_1A_head)
```


```{r}
# plot_id is id_plot
library(dplyr) # important to use dplyr, bc. it keeps the order!!!
Inn_Dike_species_1A_head_merged <- Inn_Dike_species_1A_head %>%
  left_join(biomasse, by = c("id_plot" = "plot_id"))

View(Inn_Dike_species_1A_head_merged)
```

# test! this must look exactly the same as two junks above 
```{r}
# Add grouping variable (e.g., Management type)
scores_NMDS_1A$Group_test <- Inn_Dike_species_1A_head_merged$timepoint # when header data is transposed correctly, timepoint is a col!

# Plot NMDS with ellipses for each group
ggplot(scores_NMDS_1A, aes(x = NMDS1, y = NMDS2, color = Group_test, fill = Group_test)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(type = "t", level = 0.95, alpha = 0.2, geom = "polygon") +
  theme_minimal(base_size = 14) +
  labs(title = "NMDS Plot with Group Ellipses",
       x = "NMDS1", y = "NMDS2") +
  theme(legend.position = "right")
```

```{r}
# Use the function ordisurf to plot contour lines
ordisurf(NMDS_Inn_1A,Inn_Dike_species_1A_head_merged$substrate_depth,main="",col="forestgreen")
# Finally, display species on plot
#orditorp(NMDS_Inn_1A,display="species",col="grey30",air=0.1,cex=1)
```

```{r}
dist_cov_1A <- vegdist(Inn_Dike_species_1A_ratio, method = "bray")
View(dist_cov_1A)

bd_cov_1A <- betadisper(dist_cov_1A,Inn_Dike_species_1A_head_merged$substrate_depth, type = "median")
View(bd_cov_1A)
bd_cov_1A$distances

plot(bd_cov_1A, ylab= "Bray-Curtis-Distance", xlab = "", label.cex = 0.5)
legend("topright",
       legend = levels(bd_cov_1A$group),
       col = 1:length(levels(bd_cov_1A$group)))
```
# is the distance between seeded and grown !per plot! affected by substrate depth? 
```{r}
bd_cov_1A_plots <- betadisper(dist_cov_1A,Inn_Dike_species_1A_head_merged$id_plot, type = "median")
View(bd_cov_1A_plots$group.distances)# these are the distances between the seed mix and the establishment for each plot
bd_cov_1A_plots <- as.data.frame(bd_cov_1A_plots$group.distances)

bd_cov_1A_plots <- bd_cov_1A_plots %>%
  tibble::rownames_to_column(var = "id_plot")

Inn_Dike_species_1A_head_merged <- Inn_Dike_species_1A_head_merged %>%
  left_join(bd_cov_1A_plots, by = "id_plot")

plot(Inn_Dike_species_1A_head_merged$substrate_depth,Inn_Dike_species_1A_head_merged$`bd_cov_1A_plots$group.distances`)
```


# NMDS for hay meadows 
```{r}
NMDS_Inn_22 <- metaMDS(Inn_Dike_species_22_ratio, k=2, trymax=100)
```

```{r}
stressplot(NMDS_Inn_22)
```

```{r}
NMDS_Inn_22
```

```{r}
ordiplot(NMDS_Inn_22)
```

```{r}
# Extract NMDS scores for sites
scores_NMDS_22 <- as.data.frame(scores(NMDS_Inn_22, display = "sites"))

View(Inn_Dike_species_22_head)
Inn_Dike_species_22_head <- as.data.frame(t(Inn_Dike_species_22_head))

# Add grouping variable (e.g., Management type)
scores_NMDS_22$Group <- Inn_Dike_species_22_head$timepoint

# Plot NMDS with ellipses for each group
ggplot(scores_NMDS_22, aes(x = NMDS1, y = NMDS2, color = Group, fill = Group)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(type = "t", level = 0.95, alpha = 0.2, geom = "polygon") +
  theme_minimal(base_size = 14) +
  labs(title = "NMDS Plot with Group Ellipses",
       x = "NMDS1", y = "NMDS2") +
  theme(legend.position = "right")
```

```{r}
# Add grouping variable (e.g., Management type)
scores_NMDS_22$Group2 <- interaction(Inn_Dike_species_22_head$timepoint, Inn_Dike_species_22_head$trait_constraint)

# Plot NMDS with ellipses for each group
ggplot(scores_NMDS_22, aes(x = NMDS1, y = NMDS2, color = Group2, fill = Group2)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(type = "t", level = 0.95, alpha = 0.2, geom = "polygon") +
  theme_minimal(base_size = 14) +
  labs(title = "",
       x = "NMDS1", y = "NMDS2") +
  theme(legend.position = "right")
```

## prepare and expand head; 
## tidy up todo: prepare one master header with all information; e.g. also the soil data etc. and use this from the beginning;  
```{r}
View(Inn_Dike_species_22_head)


# plot_id is id_plot
library(dplyr) # important to use dplyr, bc. it keeps the order!!!
Inn_Dike_species_22_head_merged <- Inn_Dike_species_22_head %>%
  left_join(biomasse, by = c("id_plot" = "plot_id"))

View(Inn_Dike_species_22_head_merged)
 ```



# is the distance between seeded and grown !per plot! affected by substrate depth? 
```{r}
dist_cov_22 <- vegdist(Inn_Dike_species_22_ratio, method = "bray")
View(dist_cov_22)

bd_cov_22 <- betadisper(dist_cov_22,Inn_Dike_species_22_head_merged$substrate_depth, type = "median")
plot(bd_cov_22)
```
# is the distance between seeded and grown !per plot! affected by substrate depth? 
```{r}
bd_cov_22_plots <- betadisper(dist_cov_22,Inn_Dike_species_22_head_merged$id_plot, type = "median")
View(bd_cov_22_plots$group.distances)# these are the distances between the seed mix and the establishment for each plot

bd_cov_22_plots <- as.data.frame(bd_cov_22_plots$group.distances)
colnames(bd_cov_22_plots)

bd_cov_22_plots <- bd_cov_22_plots %>%
  tibble::rownames_to_column(var = "id_plot")

Inn_Dike_species_22_head_merged <- Inn_Dike_species_22_head_merged %>%
  left_join(bd_cov_22_plots, by = "id_plot")

View(Inn_Dike_species_22_head_merged)

plot(Inn_Dike_species_22_head_merged$substrate_depth,Inn_Dike_species_22_head_merged$`bd_cov_22_plots$group.distances`)
```


# TODO: 
# is the distance between seeded and grown affected by CV of seed mixture? 
# Establishment rates and how they are affected?!

```{r}
 #load libraries
 library(tidyverse)
 library(sf)
install.packages("raster")
 library(raster)
 library(rnaturalearth)
install.packages("RefManageR")
 library(RefManageR)
```

```{r}
#Import data
load(here("data", "raw", "sPlotOpen.RData"))
 ls()
 ## [1] "CWM_CWV.oa"
 ## [4] "metadata.oa"
 "DT2.oa"
 "reference.oa"
 "header.oa"
 "sPlotOpen_citation
```






